Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-06-10 paper.Rnw
updated paper after fixing limeaid issue with the point colors

exp_scatter <- ggplot() +
  geom_point(data = sine_poi_perms,
             mapping = aes(x = x1,
                           y = x2,
                           color = rfpred,
                           size = weights),
             alpha = 0.8,
             shape = 20) +
  scale_size(range = c(0.1, 3)) +
  geom_rect(data = sine_poi_bounds %>% filter(feature == "x1"),
            mapping = aes(xmin = lower,
                          xmax = upper,
                          ymin = -Inf,
                          ymax = Inf),
            alpha = 0.25,
            fill = "grey90",
            color = "firebrick") +
  geom_rect(data = sine_poi_bounds %>%
              filter(feature == "x2"),
            mapping = aes(xmin = -Inf,
                          xmax = Inf,
                          ymin = lower,
                          ymax = upper,
                          color = coef),
            alpha = 0.25,
            fill = "grey90",
            color = "steelblue") +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1, y = x2, shape = shape, fill = rfprob_blue),
             size = 7,
             alpha = 0.8) +
  scale_shape_manual(values = 23) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("x_1"),
       y = TeX("x_2"),
       size = "Weight",
       shape = "",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       title = "Explanation Scatterplot") +
  guides(shape = guide_legend(order = 1)) +
  theme(aspect.ratio = 1)

@@

<<bullet-eoi-plots, message = FALSE>>=
bullet_explain_perms <- bullet_explain_perms %>%
  mutate(case = ifelse(case == bullet_poi_nonmatch,
                       paste(case, "(non-match)"),
                       paste(case, "(match)")))

eoi1_3qb <- plot_explain_scatter(bullet_explain_perms[1:3,], alpha = 0.9, weights = FALSE) +
  scale_color_manual(values = c("grey50", "darkorange")) +
  scale_fill_gradient2(low = "grey50",
                        high = "darkorange",
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(strip.placement = "outside",
        strip.background = element_rect(color = "white",
                                        fill = "white"))

eoi2_3qb <- plot_explain_scatter(bullet_explain_perms[4:6,], alpha = 0.9, weights = FALSE) +
  scale_color_manual(values = c("grey50", "darkorange")) +
  scale_fill_gradient2(low = "grey50",
                        high = "darkorange",
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(strip.placement = "outside",
        strip.background = element_rect(color = "white",
                                        fill = "white"))

eoi1_3eb <- plot_explain_scatter(bullet_explain_perms[7:9,], alpha = 0.9, weights = FALSE) +
  scale_color_manual(values = c("grey50", "darkorange")) +
  scale_fill_gradient2(low = "grey50",
                        high = "darkorange",
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(strip.placement = "outside",
        strip.background = element_rect(color = "white",
                                        fill = "white"))

eoi2_3eb <- plot_explain_scatter(bullet_explain_perms[10:12,], alpha = 0.9, weights = FALSE) +
  scale_color_manual(values = c("grey50", "darkorange")) +
  scale_fill_gradient2(low = "grey50",
                        high = "darkorange",
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(strip.placement = "outside",
        strip.background = element_rect(color = "white",
                                        fill = "white"))

eoi1_kd <- plot_explain_scatter(bullet_explain_perms[13:15,], alpha = 0.9, weights = FALSE) +
  scale_color_manual(values = c("grey50", "darkorange")) +
  scale_fill_gradient2(low = "grey50",
                        high = "darkorange",
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(strip.placement = "outside",
        strip.background = element_rect(color = "white",
                                        fill = "white"))

eoi2_kd <- plot_explain_scatter(bullet_explain_perms[16:18,], alpha = 0.9, weights = FALSE) +
  scale_color_manual(values = c("grey50", "darkorange")) +
  scale_fill_gradient2(low = "grey45",
                        high = "darkorange",
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(strip.placement = "outside",
        strip.background = element_rect(color = "white",
                                        fill = "white"))

@@

\begin{figure*}[!thp]
<<bullet-eois-nonmatch, out.width = '\\textwidth', fig.width = 12, fig.height = 6.25, message = FALSE>>=
# Join the plots
plot_grid(pf2_3qb, pf2_3eb, pf2_kd,
          eoi2_3qb, eoi2_3eb, eoi2_kd,
          nrow = 2,
          rel_heights = c(0.45, 0.55))
@
\caption{Plots of LIME explanations for one case in the test data that is a known non-match. Each column contains plots associated with a different application (3 quantile bins, 3 equal bins, or kernel density). The top row of plots are the explanation plots from \emph{lime}. The bottom row shows scatter plots of the simulated data associated with the prediction of interest. The features plotted are those chosen by LIME in the explanations, and the points are colored by the the random forest predictions. Lines showing the divisions created in the feature space by the bin based LIME methods are included as solid black lines. The scatter plots can be used to better understand the LIME explanation and assess if it is a good explanation.}
\label{fig:bullet-eois-nonmatch}

@@

<<bullet-eois-match, out.width = '\\textwidth', fig.width = 12, fig.height = 6.25, message = FALSE>>=
# Join the plots
plot_grid(pf1_3qb, pf1_3eb, pf1_kd,
          eoi1_3qb, eoi1_3eb, eoi1_kd,
          nrow = 2,
          rel_heights = c(0.45, 0.55))
@
\caption{Plots with the same structure as \ref{fig:bullet-eois-match} but for an observation from the test data that is a known match.}
\label{fig:bullet-eois-match}
\end{figure*}

>>>>

exp_scatter <- 
  plot_explain_scatter(sine_poi_lime_default) +
  facet_grid(switch = "both") +
  labs(x = TeX("x_1"),
       y = TeX("x_2"),
       size = "Weight",
       shape = "",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       linetype = "",
       title = "Explanation Scatterplot") +
  theme_bw(base_family = "Times", base_size = 10) +
  guides(shape = guide_legend(order = 1)) +
  theme(aspect.ratio = 1)

@@

<<bullet-eoi-plots, message = FALSE>>=
bullet_explain_perms <- bullet_explain_perms %>%
  mutate(case = ifelse(
    case == bullet_poi_nonmatch,
    paste(case, "(non-match)"),
    paste(case, "(match)")
  ))

eoi1_3qb <-
  plot_explain_scatter(bullet_explain_perms[1:3, ], alpha = 0.9, weights = TRUE) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(color = "white",
                                    fill = "white")
  ) + 
  guides(linetype = guide_legend(override.aes = list(color = c("darkorange"))))

eoi2_3qb <-
  plot_explain_scatter(bullet_explain_perms[4:6, ], alpha = 0.9, weights = TRUE) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(color = "white",
                                    fill = "white")
  ) + 
  guides(linetype = guide_legend(override.aes = list(color = c("grey50", "darkorange"))))

eoi1_3eb <-
  plot_explain_scatter(bullet_explain_perms[7:9, ], alpha = 0.9, weights = TRUE) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(color = "white",
                                    fill = "white")
  ) + 
  guides(linetype = guide_legend(override.aes = list(color = c("grey50", "darkorange"))))

eoi2_3eb <-
  plot_explain_scatter(bullet_explain_perms[10:12, ], alpha = 0.9, weights = TRUE) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(color = "white",
                                    fill = "white")
  ) + 
  guides(linetype = guide_legend(override.aes = list(color = c("grey50", "darkorange"))))

eoi1_kd <-
  plot_explain_scatter(bullet_explain_perms[13:15, ], alpha = 0.9, weights = TRUE) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(color = "white",
                                    fill = "white")
  )

eoi2_kd <-
  plot_explain_scatter(bullet_explain_perms[16:18, ], alpha = 0.9, weights = TRUE) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey45",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = "Times", base_size = 8) +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(color = "white",
                                    fill = "white")
  )

@@

\begin{figure*}[!thp]
<<bullet-eois-nonmatch, out.width = '\\textwidth', fig.width = 15, fig.height = 7.8125, message = FALSE>>=
# Join the plots
plot_grid(pf2_3qb, pf2_3eb, pf2_kd,
          eoi2_3qb, eoi2_3eb, eoi2_kd,
          nrow = 2,
          rel_heights = c(0.45, 0.55))
@
\caption{Plots of LIME explanations for one case in the test data that is a known non-match. Each column contains plots associated with a different application (3 quantile bins, 3 equal bins, or kernel density). The top row of plots are the explanation plots from \emph{lime}. The bottom row shows scatter plots of the simulated data associated with the prediction of interest. The features plotted are those chosen by LIME in the explanations, and the points are colored by the the random forest predictions. Lines showing the divisions created in the feature space by the bin based LIME methods are included as solid black lines. The scatter plots can be used to better understand the LIME explanation and assess if it is a good explanation.}
\label{fig:bullet-eois-nonmatch}

@@

<<bullet-eois-match, out.width = '\\textwidth', fig.width = 15, fig.height = 7.8125, message = FALSE>>=
# Join the plots
plot_grid(pf1_3qb, pf1_3eb, pf1_kd,
          eoi1_3qb, eoi1_3eb, eoi1_kd,
          nrow = 2,
          rel_heights = c(0.45, 0.55))
@
\caption{Plots with the same structure as \ref{fig:bullet-eois-match} but for an observation from the test data that is a known match.}
\label{fig:bullet-eois-match}
\end{figure*}
